%%%%%%
%
% PROJECT 1 - INTRO TO SCM AND LATEX
%
% filename: intro.tex
% last modified: 2014-7-22
%
%%%%%%%
%
%
%%%%%%%

\documentclass
[justified,nohyper]
{tufte-handout}

\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{kmath,kerkis} % The order of the packages matters; kmath changes the default text font
\usepackage[T1]{fontenc}

\begin{document}
\begin{center}
\end{center}
\section{Advanced Calculus Project 1: Intro to SageMathCloud and \LaTeX}
\newthought{Welcome} to the first Advanced Calculus Project!

These projects will form a large part of the course and I have selected some really exciting topics for us to study during the year. I really hope you enjoy each of them! They will challenge your abilities, but hopefully in a good way.

I'm going to assume that you are familiar with the Guidelines for Projects document that is located in the Advanced Calculus Projects folder. If you cannot find this document, please contact me either in class or by email.

The goal of this first project is to introduce you to the two main tools we will be using all year long. The first is a website called SageMathCloud (SMC) and the second is called \LaTeX$\;$ which is pronounced: ``lay''-``tech.''

There is a link to a YouTube video that I made about how to set up a SCM account and run through the first project. All you need to do is watch the video and follow the steps. Please pay attention to each step as the things you learn in this project will be used countless times throughtout the year.

Upon completion of the project, you will need to upload your final PDF report to the website turnitin.com. The video also explains how that process works.

Enjoy! 

\end{document}